Representable cylindric algebras of dimension $\omega$: the aspects of canonicity

### 2021-06-26 problem-5/Text.tex
New file with filtrations stuff
updates in cylindric notes

@@

\section{Boolean algebras with operators and cylindric algebras}

@@

\begin{defin}  Let $I$ be an index set and $\{ \Omega_i\}_{i \in I}$ a set of function symbols
\begin{enumerate}
  \item An atom structure is a relational structrure
  $\mathcal{F} = \langle W, \{ R_{i} \}_{i \in I} \rangle$
  such that $R_{i}$ is a $n+1$-ary relation symbol, if $\Omega_{i \in I}$ is an $n$-ary function symbol,
  \item Let $\mathcal{B}$ be an atomic BAO of the signature $I$,
the atom structure of $\mathcal{B}$, written as ${\bf At} \mathcal{B}$, is an atom structure $\langle \operatorname{At}(\mathcal{B}), \{ R_{i}\}_{i \in I} \rangle$ such that for each
$a, b_0, \dots, b_{n + 1} \in \operatorname{At}(\mathcal{B})$ and for each $i \in I$
\begin{center}
  ${\bf At} \mathcal{B} \models R_{i}(a,  b_0, \dots, b_{n + 1})$ iff $\mathcal{B} \models a \leq \Omega_{i}(b_0, \dots, b_{n + 1})$
\end{center}
\item Let $\mathcal{F} = \langle W, \{ R_{i} \}_{i \in I} \rangle$ be an atom structure, the complex algebra of $\mathcal{F}$, written as $\bf{Cm} \mathcal{F}$, is a BAO
$\langle \mathcal{P}(W), \cup, -, \emptyset, W, \{\Omega_{R_{i}}\}_{i \in I} \rangle$ such that
for all $X_0, \dots, X_{n - 1} \subseteq W$ and for each $i \in I$
\begin{center}
  $\Omega_{R_{i}}(X_0, \dots, X_{n - 1}) = \{ a \in W \: | \: \exists b_0 \in X_0 \dots \exists b_{n -1} \in X_{n - 1} \: \mathcal{F} \models R_{i}(a, b_0, \dots, b_{n - 1})\}$
\end{center}
\end{enumerate}
\end{defin}

@@

  \begin{enumerate}
    \item Let $\mathcal{B}$ be a complete atomic BAO, then $\mathcal{B} \cong \bf{Cm} (\bf{At}(\mathcal{B}))$,
    \item Let $\mathcal{F}$ be an atom structure, then $\mathcal{F} \cong {\bf At}({\bf Cm}(\mathcal{B}))$.
  \end{enumerate}
\end{fact}

Let $A$ be a non-empty subset of a Boolean algebra $\mathcal{B}$, $A$ is a \emph{filter}, if $A$ is closed under finite infima and upwardly closed. $A$ is an ultrafilter, if it has no non-trivial extensions. That is, if $A \subseteq A'$, then $A' = \mathcal{B}$.

@@

  \begin{enumerate}
    \item The canonical extension of $\mathcal{B}$ is a complex algebra of the canonical frame ${\bf Cm}(\mathcal{F}_{\mathcal{B}})$ denoted as $\mathcal{B}^{+}$,
    \item The class of BAOs is canonical, if it is closed under canonical extensions.
  \end{enumerate}
\end{defin}

@@

  \begin{enumerate}
    \item A subset of $\prescript{\alpha}{}U$ is an $\alpha$-ry relation on $U$. For $i, j < \alpha$, the $i,j$-diagonal $D_{ij}$ is the set of all
    elements of $U$ such that $y_i = y_j$.

    If $i < \alpha$ and $X$ is an $\alpha$-ry relation on $U$, then
    the $i$-th cylindrification $C_i X$ is the set of all elements of $U$ that agree with some element of $X$ on each coordinate except the
    $i$-th one. To be more precise,
    $C_i X = \{ y \in \prescript{\alpha}{}U \: | \: \exists x \in X \forall i < \alpha \: (i \neq j \Rightarrow y_j = x_j)\}$.
    \item A cylindic set algebra of dimension $\alpha$ is an algebra consisting of a set $S$ of $\alpha$-ry relation on some base set $U$
    with the constants and operations $0 = \emptyset$, $1 = \prescript{\alpha}{}U$, $\cap$, $-$, the diagonal elements $\{ D_{ij} \}_{i, j < \alpha}$,
    the cylindrifications $\{ C \}_{i < \alpha}$.

    A generalised cylindric set algebra of dimension $\alpha$ is a subdirect of
    cylindric algebras that have dimension $\alpha$
    \item A cylindric algebra of dimension $\alpha$ is an algebra $\mathcal{C} = \langle \mathcal{B}, \{ c_i \}_{i < \alpha}, \{ d_{ij} \}_{i, j < \alpha} \rangle$ such that
    \begin{itemize}
      \item $\mathcal{B}$ is a Boolean algebra, for each $i, j < \alpha$ $c_i$ is an operator and $d_{ij} \in \mathcal{B}$
      \item For each $i < \alpha$, $a \leq c_i a$, $c_i (a \land c_i b) = c_i a \land c_i b$ and $d_{ii} = 1$
      \item For every $i, j < \alpha$, $c_i c_j a = c_j c_i a$
      \item If $k \neq i, j < \alpha$, then $d_{ij} = c_k (d_{ij} \land d_{jk})$
      \item If $i \neq j$, then $c_i (d_{ij} \land a) \land c_i (d_{ij} \land - a) = 0$
    \end{itemize}
    ${\bf CA}_{\alpha}$ is the class of all cylindric algebras of dimension $\alpha$
    \item An $\alpha$-dimensional cylindric algebra $C$ is representable, if it is isomorphic to a generalised cylindric set algebra
    of dimension $\alpha$. Such is isomorphism is a representation of $C$.

    ${\bf RCA}_{\alpha}$ is the class of all representable cylindric algebras that have dimension $\alpha$. In particular, we are interested in the case when $\alpha = \omega$.
  \end{enumerate}
\end{defin}

One may define a representation explicitly in the following way:

@@

@@

\begin{defin} Atom
\end{defin}

Let us concretise the definition of a canonical extension for $\alpha$-dimensional cylindric algebras:

@@

\section{The result}

\begin{lemma}[See \cite{hirsch1997complete}]
  A Boolean algebra is compeletely representable iff it has an atomic representation.
\end{lemma}

@@

  $\mathcal{N}$ is a network, if the following conditions hold, for all $x, y \in \prescript{\omega}{}N$ and $i, j < \omega$:
  \begin{enumerate}
    \item $l(x) \leq d_{ij}$ iff $x(i) = x(j)$
    \item $x \equiv_i y$ implies $l(x) \leq c_i l(y)$
  \end{enumerate}
\end{defin}

@@

\begin{defin}
  Let $\mathcal{A}$ be an atomic BAO of type ${\bf CA}_{\omega}$ and $\kappa > 0$ a cardinal. The game $\mathcal{G}^{\kappa}(\mathcal{A})$ is defined as follows. The game has two players: $\forall$ (Abelard) and $\exists$ (H\'{e}lo\"{i}se). A play of the game is the sequence
  \begin{center}
    $\mathcal{N}_0 \subseteq \mathcal{N}_1 \subseteq \mathcal{N}_2 \subseteq \dots \subseteq \mathcal{N}_{\lambda} \subseteq \dots$ for $\lambda < \kappa$
  \end{center}
  The game consists of the following stages:
  \begin{enumerate}
    \item ({\bf Zero round})

@@

    Let $0 < \lambda$ be a cardinal such that $\lambda + 1 < \kappa$ and $\mathcal{N}_{\lambda} = \langle N_{\lambda}, l_{\lambda} \rangle$ is a network that has been played.

@@

  \begin{enumerate}
    \item $\Rightarrow$
    If $\mathcal{A}$ is completely representable, then its Boolean reduct is representable as well. $\exists$ maintains that embedding to win the play.

@@

    Suppose $\exists$ has a winning strategy in $\mathcal{G}^{\kappa + \omega}(\mathcal{A})$.
    In every round $\forall$ picks all possible $i < \omega$, $a \in \operatorname{At}(\mathcal{A})$, all possible hyperedges and all appropriate atoms.
    ...
    Question: Is this claim correct?
  \end{enumerate}
\end{proof}

@@

\section{Sahlqvist and canonical axiomatisation of ${\bf CA}_{\omega}$}

>>>>

\let\strokeL\L
\DeclareRobustCommand{\L}{\ifmmode\mathbf{L}\else\strokeL\fi}

@@

\section{Boolean algebras with operators and cylindric algebras, a bit of the backgroud}

Let $a \in \mathcal{B}$ be an element of a Boolean algebra $\mathcal{B}$, $a$ is called an atom, if for every $b \in \mathcal{B}$
$b < a$ implies $b = 0$. That is, an atom is a minimal non-zero element. $\operatorname{At}(\mathcal{B})$ is the set of all atoms
of $\mathcal{B}$.

Let $\mathcal{B}$ be a Boolean algebra and $\mathcal{F}$ a field of sets such that $h : \mathcal{B} \to \mathcal{F}$ is a
representation of $\mathcal{B}$, then $\mathcal{B}$ is a complete representation of $\mathcal{B}$, if for every
$A \subseteq \mathcal{B}$ whenever $\Sigma A$ we have the following:
\begin{center}
  $h(\Sigma A) = \bigcup h[A]$
\end{center}

\begin{theorem} \label{completeboolean}
  Let $\mathcal{B}$ be a Boolean algebra, then $\mathcal{B}$ is atomic iff $\mathcal{B}$ is completely representable. See \cite[Corollary 6]{hirsch1997complete}.
\end{theorem}

@@

\begin{defin}  Let $I$ be an index set and $\{ \Omega_i\}_{i \in I}$ a set of function symbols
\begin{enumerate}
  \item An atom structure is a relational structrure
  $\mathcal{F} = \langle W, \{ R_{i} \}_{i \in I} \rangle$
  such that $R_{i}$ is a $n+1$-ary relation symbol, where $\Omega_{i \in I}$ is an $n$-ary function symbol,
  \item Let $\mathcal{B}$ be an atomic BAO of the signature $I$,
the atom structure of $\mathcal{B}$, written as $\mathfrak{At} \mathcal{B}$, is an atom structure $\langle \operatorname{At}(\mathcal{B}), \{ R_{i}\}_{i \in I} \rangle$ such that for each
$a, b_0, \dots, b_{n + 1} \in \operatorname{At}(\mathcal{B})$ and for each $i \in I$
\begin{center}
  $\mathfrak{At} \mathcal{B} \models R_{i}(a,  b_0, \dots, b_{n + 1})$ iff $\mathcal{B} \models a \leq \Omega_{i}(b_0, \dots, b_{n + 1})$
\end{center}
\item Let $\mathcal{F} = \langle W, \{ R_{i} \}_{i \in I} \rangle$ be an atom structure, the complex algebra of $\mathcal{F}$, written as $\bf{Cm} \mathcal{F}$, is a BAO
$\langle \mathcal{P}(W), \cup, -, \emptyset, W, \{\Omega_{R_{i}}\}_{i \in I} \rangle$ such that
for all $X_0, \dots, X_{n - 1} \subseteq W$ and for each $i \in I$
\begin{center}
  $\Omega_{R_{i}}(X_0, \dots, X_{n - 1}) = \{ a \in W \: | \: \exists b_0 \in X_0 \dots \exists b_{n -1} \in X_{n - 1} \: \mathcal{F} \models R_{i}(a, b_0, \dots, b_{n - 1})\}$
\end{center}
\end{enumerate}
\end{defin}

@@

  \begin{enumerate}
    \item Let $\mathcal{B}$ be a complete atomic BAO, then $\mathcal{B} \cong \mathfrak{Cm} (\bf{At}(\mathcal{B}))$,
    \item Let $\mathcal{F}$ be an atom structure, then $\mathcal{F} \cong \mathfrak{At}(\mathfrak{Cm}(\mathcal{B}))$.
  \end{enumerate}
\end{fact}

Let $A$ be a non-empty subset of a Boolean algebra $\mathcal{B}$, $A$ is a \emph{filter}, if $A$ is closed under finite infima and upward closed. $A$ is an ultrafilter, if it has no non-trivial extensions. That is, if $A \subseteq A'$, then $A' = \mathcal{B}$. This is a well-known fact that every filter can be extended to a maximal one using Zorn's lemma.

The following definition is due to, for example, \cite[Definition 5.40]{venema2010}.

@@

  \begin{enumerate}
    \item The canonical extension of $\mathcal{B}$ is a complex algebra of the canonical frame $\mathfrak{Cm}(\mathcal{F}_{\mathcal{B}})$ denoted as $\mathcal{B}^{+}$,
    \item The class of BAOs is canonical, if it is closed under canonical extensions.
  \end{enumerate}
\end{defin}

@@

A subset of $\prescript{\alpha}{}U$ is an $\alpha$-ry relation on $U$. For $i, j < \alpha$, the \emph{$i,j$-diagonal} $D_{ij}$ is the set of all elements of $\prescript{\alpha}{}U$ such that $y_i = y_j$.

If $i < \alpha$ and $X$ is an $\alpha$-ry relation on $U$, then
the $i$-th cylindrification $C_i X$ is the set of all elements of $U$ that agree with some element of $X$ on each coordinate except, perhaps, the $i$-th one. To be more precise,
\begin{center}
$C_i X = \{ y \in \prescript{\alpha}{}U \: |
\: \exists x \in X \forall i < \alpha \: (i \neq j \Rightarrow y_j = x_j)\}$.
\end{center}
We define the following equivalence relation for $i < \alpha$ and $x, y \in \prescript{\alpha}{}U$:
\begin{center}
  $x \equiv_i y \Leftrightarrow \forall j \in \alpha \: (i \neq j \Rightarrow x(i) = y(j))$
\end{center}
Then one may reformulate the definition of the $i$-th cylindrification in the following way:

\begin{center}
  $C_i X = \{ y \in \prescript{\alpha}{}U \: | \: \exists x \in X \:\: x \equiv_i y \}$
\end{center}

According to this version of the definiton, one may think of the cylindrification as an ${\bf S}5$ modal operator.

The following definition is due to \cite{sankappanavar1981course}:

\begin{defin}
  Let $(\mathcal{A}_i)_{i \in I}$ be a family of algebras (of an abstract signature) and $\mathcal{A}$ is a subalgebra of $\prod \limits_{i \in I} \mathcal{A}_i$, then $\mathcal{A}$ is a subdirect product, if every projection is onto. That is, for every $i \in I$, $\pi_i[\mathcal{A}] = \mathcal{A}_i$.
\end{defin}

\begin{defin}
  A cylindic set algebra of dimension $\alpha$ is an algebra consisting of a set $S$ of $\alpha$-ry relation on some base set $U$
    with the constants and operations $0 = \emptyset$, $1 = \prescript{\alpha}{}U$, $\cap$, $-$, the diagonal elements $\{ D_{ij} \}_{i, j < \alpha}$, the cylindrifications $\{ C_i \}_{i < \alpha}$. A generalised cylindric set algebra of dimension $\alpha$ is a subdirect of cylindric algebras that have dimension $\alpha$
\end{defin}

\begin{defin}
    A cylindric algebra of dimension $\alpha$ is an algebra $\mathcal{C} = \langle \mathcal{B}, \{ c_i \}_{i < \alpha}, \{ d_{ij} \}_{i, j < \alpha} \rangle$ such that
    \begin{itemize}
      \item $\mathcal{B}$ is a Boolean algebra, for each $i, j < \alpha$ $c_i$ is an operator and $d_{ij} \in \mathcal{B}$
      \item For each $i < \alpha$, $a \leq c_i a$, $c_i (a \land c_i b) = c_i a \land c_i b$ and $d_{ii} = 1$
      \item For every $i, j < \alpha$, $c_i c_j a = c_j c_i a$
      \item If $k \neq i, j < \alpha$, then $d_{ij} = c_k (d_{ij} \land d_{jk})$
      \item If $i \neq j$, then $c_i (d_{ij} \land a) \land c_i (d_{ij} \land - a) = 0$
    \end{itemize}
    ${\bf CA}_{\alpha}$ is the class of all cylindric algebras of dimension $\alpha$
\end{defin}

One may define a representation of a cylindric algebra explicitly in the following way:

@@

An $\alpha$-dimensional cylindric algebra $C$ is representable, if it is isomorphic to a generalised cylindric set algebra
of dimension $\alpha$. Such is isomorphism is a representation of $C$. ${\bf RCA}_{\alpha}$ is the class of all representable cylindric algebras that have dimension $\alpha$. In particular, we are interested in the case when $\alpha = \omega$.

@@

Let us concretise the definition of a canonical extension for ${\bf CA}_{\alpha}$-type BAOs.

@@

\section{Completely representable cylindric algebras of dimension $\omega$}

@@

  $\mathcal{N}$ is a network, if the following conditions hold, for all $x, y \in \prescript{\omega}{}N$ and $i, j < \omega$:
  \begin{enumerate}
    \item $l(x) \leq d_{ij}$ iff $x_i = x_j$
    \item $x \equiv_i y$ implies $l(x) \leq c_i l(y)$
  \end{enumerate}
\end{defin}

@@

\begin{defin}
  Let $\mathcal{A}$ be an atomic BAO of type ${\bf CA}_{\omega}$ and $\kappa > 0$ a cardinal. The game $\mathcal{G}^{\kappa}(\mathcal{A})$ is defined as follows. The game has two players: $\forall$ (Abelard, he/his) and $\exists$ (H\'{e}lo\"{i}se, she/her). A play of the game $\mathcal{G}^{\kappa}(\mathcal{A})$ is the sequence of networks
  \begin{center}
    $\mathcal{N}_0 \subseteq \mathcal{N}_1 \subseteq \mathcal{N}_2 \subseteq \dots \subseteq \mathcal{N}_{\lambda} \subseteq \dots$ for $\lambda < \kappa$
  \end{center}
  The game consists of the following stages:
  \begin{enumerate}
    \item ({\bf Zero round})

@@

    Let $0 < \lambda$ be a cardinal such that $\lambda + 1 < \kappa$ and a network $\mathcal{N}_{\lambda} = \langle N_{\lambda}, l_{\lambda} \rangle$ has been already played.

@@

  \begin{enumerate}
    \item $\Rightarrow$
    If $\mathcal{A}$ is completely representable, then its Boolean reduct is completely representable as well by Theorem~\ref{completeboolean}. $\exists$ maintains that embedding to win the play. TODO: write down this proof in more detail

@@

    Suppose $\exists$ has a winning strategy in $\mathcal{G}^{\kappa + \omega}(\mathcal{A})$.
    In every round $\forall$ picks all possible $i < \omega$, $a \in \operatorname{At}(\mathcal{A})$, all possible hyperedges and all appropriate atoms and $\exists$ has a proper response for every $\forall$'s move.

    For each atom consider a play of the game with fewer than $\kappa + \omega$ nodes. For each $a \in \operatorname{At}(\mathcal{A})$ we associate a network $\mathcal{N}_a$, the resulting network of a corresponding game. Consider the set $\{ \mathcal{N}_a \: | \: a \in \operatorname{At}(\mathcal{A}) \}$.

    Let $a$ be an atom, consider the network $\mathcal{N}_a = \langle V, l_a \rangle$. If there was not $x \in V$ such that $l_a(x) = a$, then $\forall$ would have a winning strategy, but that is not true, such an $x$ does exist. The second item of this criterion follows from the presence of a winning strategy for $\exists$ as well.

    So we define a map $rep$:
    \begin{center}
    $rep(a) = \{ x \: | \: \exists b \in \operatorname{At}(\mathcal{A}) \: x \in \prescript{\omega}{}N_a, l_a(x) \leq b \}$.
  \end{center}
    We check that $rep$ preserves cylindrifications and diagonal elements. Let $i, j < \omega$ and $a \in \mathcal{A}$:
    \begin{enumerate}
    \item Suppose $x \in rep(c_i a)$, then there exists an atom $b$ such that $x \in \prescript{\omega}{}N_b$ with $l_b(x) \leq c_i a$. Then there exists $y \equiv_i x$ with $l_b(y) \leq a$, so $x \in {\bf C}_i(rep(a))$.

    If $x \in {\bf C}_i(rep(a))$, then there exists $y \equiv_i x$ such that $y \in rep(a)$, that is, there exists an atom $b$ such that $y \in \prescript{\omega}{}N_b$ and $l_b(y) \leq a$.
    \item If $x \in rep(d_{ij})$, so there exists an atom $b$ with $x \in \prescript{\omega}{}N_a$ and $l_b(x) \leq d_{ij}$, then $x_i = x_j$, then $x \in D_{ij}$.
    \end{enumerate}
  \end{enumerate}
\end{proof}

\begin{theorem} Let $\mathcal{A}$ be a BAO of type ${\bf CA}_{\omega}$:

\begin{enumerate}
  \item $\exists$ has a winning strategy in $\mathcal{G}_{m}(\mathcal{A})$ ($m < \omega$), then $\exists$ has a winning strategy in $\mathcal{G}_{\omega}(\Pi_U \mathcal{A})$, where $\Pi_U \mathcal{A}$ is the non-principal ultrapower of $\mathcal{A}$ modulo ultrafilter $U$ over $\omega$.
  \item $\exists$ has a winning strategy in $\mathcal{G}_{m}(\mathcal{A})$ (for every $m < \omega$) iff $\mathcal{A}$ is elementarily equivalent to a completely representable cylindric algebra of dimension $\omega$.
\end{enumerate}
\end{theorem}

  The argument uses \L o\'s's Theorem, see \cite[Theorem 8.5.3]{hodges1997shorter}.

  \begin{enumerate}
    \item
    \item
  \end{enumerate}
\end{proof}

\section{The result itself}

@@

\section{(Lack of) canonical axiomatisation of ${\bf CA}_{\omega}$}

Here we are going to show that ${\bf CA}_{\omega}$ fails to have a canonical axiomatisation, the similar results for ${\bf RRA}$ and ${\bf RCA}_n$ for finite $n \geq 3$ have been shown by Hodkinson and Venema \cite{hodkinson2005canonical} and by Bulian and Hodkinson respectively \cite{bulian2013bare}.
